Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm':' A Case Study

### 2022-04-14 Paper_Draft2/Draft2.Rmd
trying to fix images

The data we are using for this project is the COMPAS General Recidivism Risk Scores data set from the AI Fairness 360 (AIF360) toolkit [@aif360-oct-2018], which does the same initial pre-processing as ProPublica. The raw data has 6,167 rows and each row represents an arrest charge for a defendant. AIF360’s COMPAS data includes the defendant's age, race, sex, what they were charged with, and whether or not the defendant ultimately recidivated within a two-year period after their arrest. For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American. Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race, age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot}).

```{r table snip, echo = FALSE, fig.align='center', fig.cap= 'This table is a snippet the data set we will be using.', out.width= "400px", out.height= "200px"}
knitr::include_graphics("../images/table_snippet.png")
```

```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.', out.width= "400px", out.height= "200px"}
knitr::include_graphics("../images/race_bar_plot.png")
```

```{r age plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant ages by race.', out.width= "400px", out.height= "200px"}
knitr::include_graphics("../images/race_age_plot.png")
```

```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.', out.width= "400px", out.height= "200px"}
knitr::include_graphics("../images/recid_bar_plot.png")
```

```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.', out.width= "400px", out.height= "200px"}
knitr::include_graphics("../images/race_recid_plot.png")
```

@@

\\textbf{Statistical Parity Difference} - This metric measures the difference that the majority and protected classes get a particular outcome. The ideal value of this metric is 0. Fairness for this metric is between -0.1 and 0.1. A negative value means there is higher benefit for the privileged group (in this case, white defendants).

@@

\\textbf{Disparate Impact Ratio} - This metric is the ratio of how often the favorable outcome occurs in one group versus the other. In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many black defendants are predicted to not recidivate. A value of 1 means that the ratio is exactly 1:1. Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits. According to AIF360, a ratio between .8 to 1.25 is considered fair:

@@

\\textbf{Average Odds Difference} - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups. A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group. Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants.

@@

\\textbf{Equal Opportunity Difference} - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups. The true positive rate is the ratio of true positives to the total number of actual positives for a given group.

@@

To reiterate, fairness is subjective, so the questions we ask ourselves to create parameters for our definition of fairness include:

>>>>

The data we are using for this project is the COMPAS General Recidivism Risk Scores data set from the AI Fairness 360 (AIF360) toolkit [@aif360-oct-2018], which does the same initial pre-processing as ProPublica. The raw data has 6,167 rows and each row represents an arrest charge for a defendant. AIF360’s COMPAS data includes the defendant's age, race, sex, what they were charged with, and whether or not the defendant ultimately recidivated within a two-year period after their arrest. For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American. Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot}).

```{r table snip, echo = FALSE, fig.align='center', fig.cap= 'This table is a snippet of the data set we will be using.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/table_snippet_new.png")
```

```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/race_bar_plot.png")
```

```{r age plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant ages by race.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/race_age_plot.png")
```

```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/recid_bar_plot.png")
```

```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/race_recid_plot.png")
```

@@

**Statistical Parity Difference** - This metric measures the difference that the majority and protected classes get a particular outcome. The ideal value of this metric is 0. Fairness for this metric is between -0.1 and 0.1. A negative value means there is higher benefit for the privileged group (in this case, white defendants).

@@

**Disparate Impact Ratio** - This metric is the ratio of how often the favorable outcome occurs in one group versus the other. In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many black defendants are predicted to not recidivate. A value of 1 means that the ratio is exactly 1:1. Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits. According to AIF360, a ratio between .8 to 1.25 is considered fair:

@@

**Average Odds Difference** - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups. A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group. Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants.

@@

**Equal Opportunity Difference** - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups. The true positive rate is the ratio of true positives to the total number of actual positives for a given group.

@@

### 2022-04-14 Paper_Draft2/Draft2.tex
trying to fix images

The data we are using for this project is the COMPAS General Recidivism
Risk Scores data set from the AI Fairness 360 (AIF360) toolkit
\citep{aif360-oct-2018}, which does the same initial pre-processing as
ProPublica. The raw data has 6,167 rows and each row represents an
arrest charge for a defendant. AIF360's COMPAS data includes the
defendant's age, race, sex, what they were charged with, and whether or
not the defendant ultimately recidivated within a two-year period after
their arrest. For the purposes of our project, which endeavors to
evaluate the differing effects of the COMPAS algorithm on white
defendants versus Black defendants, we have filtered the data to only
include individuals whose race is listed as Caucasian or
African-American. Our data therefore has 5,723 rows (Figure
\ref{fig:table snip}), with the below distributions of race, age (Figure
\ref{fig:age plot}), and two year recidivism rate (Figures
\ref{fig:recid plot} \& \ref{fig:recid race plot}).

@@

{\centering \includegraphics[width=400px,height=200px]{../images/table_snippet}

@@

\caption{This table is a snippet the data set we will be using.}\label{fig:table snip}
\end{figure}

@@

{\centering \includegraphics[width=400px,height=200px]{../images/race_bar_plot}

@@

{\centering \includegraphics[width=400px,height=200px]{../images/race_age_plot}

@@

{\centering \includegraphics[width=400px,height=200px]{../images/recid_bar_plot}

@@

{\centering \includegraphics[width=400px,height=200px]{../images/race_recid_plot}

@@

\textbackslash textbf\{Statistical Parity Difference\} - This metric
measures the difference that the majority and protected classes get a
particular outcome. The ideal value of this metric is 0. Fairness for
this metric is between -0.1 and 0.1. A negative value means there is
higher benefit for the privileged group (in this case, white
defendants).

@@

\textbackslash textbf\{Disparate Impact Ratio\} - This metric is the
ratio of how often the favorable outcome occurs in one group versus the
other. In the case of recidivism, this is the ratio of how many white
defendants are predicted to not recidivate compared to how many black
defendants are predicted to not recidivate. A value of 1 means that the
ratio is exactly 1:1. Less than 1 means the privileged group (white
defendants) benefits, while a value greater than 1 means the
unprivileged group (Black defendants) benefits. According to AIF360, a
ratio between .8 to 1.25 is considered fair:

@@

\textbackslash textbf\{Average Odds Difference\} - This metric returns
the average difference in false positive rate and true positive rate for
the privileged and unprivileged groups. A value of 0 indicates equality
of odds, and a value below 0 implies benefit for the privileged group.
Equality of odds is achieved in the case of recidivism when the
proportion of people who were predicted to recidivate and did recidivate
is equal (true positive rate) for both Black and white defendants AND
the proportion of people who were predicted to recidivate and did not
recidivate (false positive rate) is equal for both Black and white
defendants.

@@

\textbackslash textbf\{Equal Opportunity Difference\} - This metric is
computed as the difference of true positive rates between the
unprivileged and the privileged groups. The true positive rate is the
ratio of true positives to the total number of actual positives for a
given group.

>>>>

The data we are using for this project is the COMPAS General Recidivism
Risk Scores data set from the AI Fairness 360 (AIF360) toolkit
\citep{aif360-oct-2018}, which does the same initial pre-processing as
ProPublica. The raw data has 6,167 rows and each row represents an
arrest charge for a defendant. AIF360's COMPAS data includes the
defendant's age, race, sex, what they were charged with, and whether or
not the defendant ultimately recidivated within a two-year period after
their arrest. For the purposes of our project, which endeavors to
evaluate the differing effects of the COMPAS algorithm on white
defendants versus Black defendants, we have filtered the data to only
include individuals whose race is listed as Caucasian or
African-American. Our data therefore has 5,723 rows (Figure
\ref{fig:table snip}), with the below distributions of race (Figure
\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year
recidivism rate (Figures \ref{fig:recid plot} \&
\ref{fig:recid race plot}).

@@

{\centering \includegraphics[width=500px,height=300px]{../images/table_snippet_new}

@@

\caption{This table is a snippet of the data set we will be using.}\label{fig:table snip}
\end{figure}

@@

{\centering \includegraphics[width=500px,height=300px]{../images/race_bar_plot}

@@

{\centering \includegraphics[width=500px,height=300px]{../images/race_age_plot}

@@

{\centering \includegraphics[width=500px,height=300px]{../images/recid_bar_plot}

@@

{\centering \includegraphics[width=500px,height=300px]{../images/race_recid_plot}

@@

\textbf{Statistical Parity Difference} - This metric measures the
difference that the majority and protected classes get a particular
outcome. The ideal value of this metric is 0. Fairness for this metric
is between -0.1 and 0.1. A negative value means there is higher benefit
for the privileged group (in this case, white defendants).

@@

\textbf{Disparate Impact Ratio} - This metric is the ratio of how often
the favorable outcome occurs in one group versus the other. In the case
of recidivism, this is the ratio of how many white defendants are
predicted to not recidivate compared to how many black defendants are
predicted to not recidivate. A value of 1 means that the ratio is
exactly 1:1. Less than 1 means the privileged group (white defendants)
benefits, while a value greater than 1 means the unprivileged group
(Black defendants) benefits. According to AIF360, a ratio between .8 to
1.25 is considered fair:

@@

\textbf{Average Odds Difference} - This metric returns the average
difference in false positive rate and true positive rate for the
privileged and unprivileged groups. A value of 0 indicates equality of
odds, and a value below 0 implies benefit for the privileged group.
Equality of odds is achieved in the case of recidivism when the
proportion of people who were predicted to recidivate and did recidivate
is equal (true positive rate) for both Black and white defendants AND
the proportion of people who were predicted to recidivate and did not
recidivate (false positive rate) is equal for both Black and white
defendants.

@@

\textbf{Equal Opportunity Difference} - This metric is computed as the
difference of true positive rates between the unprivileged and the
privileged groups. The true positive rate is the ratio of true positives
to the total number of actual positives for a given group.
